Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections

### 2022-02-27 _paper.tex
text changes from thesis, Needed: references and figure changes

\begin{document}
\maketitle
\begin{abstract}
The increased predictive power comes at the cost of interpretability, which has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models use predictors to arrive at a prediction with a point estimate of the linear feature importance in the vicinity of each instance. These can be considered linear projections and can be further explored interactively to understand better the interaction between features used to make predictions across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any instance and illustrate with examples with categorical (penguin species, chocolate types) and quantitative (football salaries, house prices) response features. The methods are implemented in the \textbf{R} package \textbf{cheem}, available on CRAN.
\end{abstract}

@@

The remainder of this paper is organized as follows. The following Section, \ref{sec:explanations}, covers the background of the local explanation and the traditional visuals produced. Section \ref{sec:tour} explains the animations of continuous linear projections. Section \ref{sec:cheemviwer} discusses the visual layout in the interactive interface, how they facilitate analysis, data preprocessing, and package infrastructure. Then Section \ref{sec:casestudies} illustrates the application to supervised learning with categorical and quantitative response features. We conclude with Section \ref{sec:cheemdiscussion} of the insights gained and directions that might be explored in the future.

@@

SHaply Additive exPlanations (SHAP) quantifies the feature contributions of one instance by examining the effect of other features on the predictions. The explanations of SHAP almost all refer to Shapley (1953)'s method to evaluate an individual's contribution to cooperative games by assessing the performance of this player in the presence or absence of other players. Strumbelj and Kononenko (2010) introduced the use of SHAP for local explanations in ML models. The attribution of feature importance depends on the sequence of the included features. The SHAP values are the mean contributions over different feature sequences. The approach is related to partial dependence plots (Molnar 2020), used to explain the effect of a feature by predicting the response for a range of values on this feature after fixing the value of all other features to their mean. Partial dependence plots are a global approximation of the feature importance, while SHAP is specific to one instance. It could also be considered similar to examining the coefficients from all subsets regression, as described in Wickham, Cook, and Hofmann (2015), which helps to understand the relative importance of each feature in the context of all other candidate features.

@@

\caption{Illustration SHAP values for a random forest model for salaries of FIFA 2020 players based on nine predictors corresponding to different skills. A star offensive and defensive player are compared, L. Messi and V. van Dijk, respectively. Panel (a) shows breakdown plots of three sequences of the features, order, and magnitude change. Panel (b) shows the distribution of attribution for each feature across 25 sequences of predictors, with the mean displayed as a dot for each player. Offense and movement are important for Messi but not van Dijk, and conversely, defense and power are important for van Dijk but not Messi.}\label{fig:shapdistrbd}
\end{figure}

For our application, we use \emph{tree SHAP}, a variant of SHAP that enjoys a lower computational complexity (Lundberg, Erion, and Lee 2018). Instead of aggregating over sequences of the features, tree SHAP calculates instance-level feature importance by exploring the structure of the decision trees. Tree SHAP is only compatible with tree-based models; we illustrate random forests. The following section will use normalized SHAP values as a projection basis (call this the \emph{attribution projection}) will have coefficients varied to further scrutinize the feature contributions.

Following the use case \emph{Explanatory Model Analysis} (Biecek and Burzykowski 2021), we use FIFA data to illustrate SHAP use. Consider soccer data from the FIFA 2020 season (Leone 2020). There are 5000 instances of 9 skill measures (after aggregating highly correlated features). A random forest model is fit regressing wages {[}2020 Euros{]} from the skill measures. We then extract the SHAP values of a star offensive player (L. Messi) and defensive player (V. van Dijk). The results are displayed in Figure \ref{fig:shapdistrbd}. We expect to see a difference in the attribution of the feature importance across the two positions of the players, which would be interpreted as how the player's salary depends on this combination of skill sets. Plot (b) is a modified breakdown plot (Gosiewska and Biecek 2019) where the order of features is fixed, so the two instances can be more easily compared.

In summary, these plots highlight how local explanations bring interpretability to a model, at least in the vicinity of their instances. In this instance, two players with different positions receive different profiles of feature importance to explain the prediction of their wages.

@@

To explore the local explanations, an ensemble of plots (Unwin and Valero-Mora 2018) is provided, called the \emph{cheem viewer}. There are two primary plots: the \textbf{global view} to give the context of all of the SHAP values, and the \textbf{radial tour view} to explore the local explanations with user-controlled rotation. In addition, there are numerous user inputs, including feature selection for the radial tour and instance selection for making comparisons. There are different plots used for the categorical and quantitative responses. Figures \ref{fig:classificationcase} and \ref{fig:regressioncase} are screenshots showing the cheem viewer for the two primary tasks: classification (categorical response) and regression (quantitative response).

@@

The global view provides the context of all instances and facilitates the exploration of the separability of the data- and attribution-spaces. Both of these spaces are of dimension \(n\times p\), where \(n\) is the number of instances and \(p\) is the number of predictors. The attribution space corresponds to the local explanations for each instance, each a vector of \(p\) values.

A visualization is provided by the first two principal components of the data (left) and the attribution (middle) spaces. These single 2D projections will not reveal all of the structure of higher-dimensional space, but they are useful visual summaries. In addition, a plot of the observed against predicted response values is also provided (Figures \ref{fig:classificationcase}b, \ref{fig:regressioncase}a), to help identify instances poorly predicted by the model. For classification tasks, misclassified instances are circled in red if applicable. Linked brushing between the plots is provided, and a tabular display of selected points helps to facilitate exploration of the spaces and the model (shown in Figures \ref{fig:classificationcase}a,c).

While the comparison of these spaces is interesting, a main purpose of the global view is to enable the selection of instances to explore the local explanations. The projection attribution of the primary instance (PI) is examined and typically viewed with an optional comparison instance (CI). These instances are highlighted as asterisk and \(\times\), respectively.

@@

The local explanations for all observations are normalized (squared sum of values adds to 1), and thus, the relative importance of features can be compared across all instances. These are depicted as a vertical parallel coordinate plot, where each line connects one instance's feature attribution (Figures \ref{fig:classificationcase}e and \ref{fig:regressioncase}e). The attribution projections of the PI and CI are shown as dashed and dotted lines, respectively. From this plot, we would read the range of importances across all instances can be interpreted. For classification, one would look at differences between groups on any feature. For example, Figure \ref{fig:classificationcase}e suggests that \texttt{b\_l} is important for distinguishing the green class from the other two. For regression, one might generally observe which features have low values for all instances (not important), for example, \texttt{BMI} and \texttt{pwr} in Figure \ref{fig:regressioncase}e, and which have a range of high and low values (e.g.~\texttt{off}, \texttt{def}) suggesting important for some instances and not important for other instances.

The overlaid bars on the parallel coordinate plot represent the attribution projection of the PI. (Remember that the PI is interactively selected from the global view). The attribution projection is an approximation of the feature importance for the prediction of this instance. This combination of features best explains the difference between the mean response and an instance's predicted value. It is not an indication of the local shape of the model surface. That is, it is not some indication of the tangent to the curve at this point.

The attribution projection of the PI is the initial 1D basis in a radial tour, displayed as a density plot for a categorical response (Figure \ref{fig:classificationcase}f), and as scatterplots for a quantitative response (Figure \ref{fig:regressioncase}f). The PI and CI are indicated by vertical dashed and dotted lines, respectively. The user uses the radial tour to vary the contribution of the selected feature between 0-1. Doing so tests the sensitivity of structure (class separation or strength of relationship) to the feature's contribution. For classification, if the separation between classes diminishes when the feature contribution is reduced then this suggests that the feature is important for the class separation. For regression, if the relationship scatterplot weakens when the feature contribution is reduced then this suggests that the feature is important for accurately predicting the response.

@@

Selecting a misclassified instance as PI and a correctly classified point nearby in data space as CI makes it easier to examine the features most responsible for the error. The global view (Figure \ref{fig:classificationcase}c) displays the model confusion matrix. The radial tour is 1D, plotted as a density where color indicates class. A scroll bar here enables the user to vary the contribution of each feature to explore the sensitivity of the separation to that feature.

@@

\caption{Overview of the cheem viewer for classification tasks. Global view inputs, (a), set the PI, CI, and color statistic. Global view, (b) PC1 by PC2 approximations of the data space and attribution space. (c) prediction by observed y (visual of the confusion matrix for classification tasks). Points are colored by predicted class, and red circles indicate misclassified instances. Radial tour inputs (d) select features to include and which feature is changed in the tour. (e) shows a parallel coordinate display of the distribution of the feature attributions while bars depict contribution for the current basis. The black bar is the variable being changed in the radial tour. Panel (f) is the resulting projection of the data indicated as density in the classification case.}\label{fig:classificationcase}
\end{figure}

@@

Selecting an inaccurately predicted instance as PI and an accurately predicted instance, with similar feature values, as CI is a useful way to understand how the model is failing or not. The global view (Figure \ref{fig:regressioncase}a) shows a scatterplot of the observed vs predicted values, which should exhibit a strong relationship if the model is a good fit. The points can be colored by a statistic, residual, a measure of outlyingness (log Mahalanobis distance), or correlation, to help with understand where the model fits better or worse.

In the radial tour view, the observed response and the residuals (vertical) are plotted against the attribution projection of the PI (horizontal). The attribution projection can be interpreted similarly to the predicted value from the global view plot. It represents a linear combination of the features, and a good fit would be indicated when there is a strong relationship seen with the observed values. This can be viewed as a local linear approximation if the fitted model is nonlinear. As the contribution of a feature is varied, if the value of the PI does not change much, it would indicate that the prediction for this instance is NOT sensitive to that feature. Conversely, if the predicted value varies substantially, the prediction is very sensitive to that feature, suggesting that the feature is very important for the PI's prediction.

@@

\caption{Overview of the cheem viewer for regression task highlighting the differences from the classification task and interactive features. Panel (a) PCA of the data and attributions spaces, (b), residual plot, predictions by observed values. Four points are selected points and highlighted in the PC spaces and tabularly displayed. Coloring on a statistic (c) highlights structure organized in the attribution space. Interactive tabular display (d) populates when instances are selected. Contribution of the 1D basis affecting the horizontal position (e) parallel coordinate display of the feature attribution from all observations, and horizontal bars show the contribution to the current basis. Regression projection (f) uses the same horizontal projection and fixes the vertical positions to the observed y and residuals (left and right).}\label{fig:regressioncase}
\end{figure}

@@

The application has several reactive inputs that affect the data used, aesthetic display, and tour manipulation. These reactive inputs make the software flexible and extensible. The application also has more exploratory interactions to help link points across displays and reveal structure found in different spaces.

A tooltip displays instance number/name and classification information while the cursor hovers over a point. Linked brushing allows the selection of points (left click and drag) where those points will be highlighted across plots. The information corresponding to the selected points is populated on a dynamic table. These interactions aid exploration of the spaces and, finally, identification of a primary and comparison instance.

@@

It is vital to mitigate the render time of visuals, especially when users may want to iterate many times. All computational operations should be prepared before runtime. The work remaining when an application is run solely reacts to inputs and rendering visuals and tables. Below we discuss the steps and details of the reprocessing.

\begin{itemize}
    \item \textbf{Data:} predictors and response are unscaled complete numerical matrix. Most models and local explanations are scale-invariant.
    \item \textbf{Model and explanation:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} (Liaw and Wiener 2002), compatibility tree SHAP. We use modest hyperparameters, namely: 125 trees, number features randomly sampled at each split, mtry = $\sqrt{p}$ or $p/3$ for classification and regression, and minimum size of terminal nodes $max(1, n/500)$ or $max(5, n/500)$ for classification and regression. Tree SHAP is calculated for \emph{each} instance using the package \textbf{treeshap} (Kominsarczyk et al. 2021). This implementation aggregates exhaustively over all trees' attribution, and we opt not to fit interactions of features. 
    \item \textbf{Cheem view:} after the model and full explanation space are calculated, we scale each feature by standard deviations away from the mean to achieve common support for visuals. Statistics for mapping to color are computed on the scaled spaces. Interactive tabular display reports the original values.
\end{itemize}

The time to preprocess the data will vary significantly with the model and local explanation. For reference, the FIFA data, 5000 instances of nine explanatory features, took 2.5 seconds to fit a random forest model of modest hyperparameters. Extracting the tree SHAP values of each instance took 270 seconds combined. PCA and statistics of the features and attributions took 2.8 seconds. These runtimes were from a non-parallelized R session on a modern laptop, but suffice to say that most of the time will be spent on the local attribution. An increase in model complexity or data dimensionality will quickly become an obstacle. Its reduced computational complexity makes tree SHAP a good candidate to start. (Alternatively, the package \textbf{fastshap} (Greenwell 2020) claims extremely low runtimes, attributed to fewer calls to the prediction function, partial implementation in C++, and efficient use of logical subsetting.)

@@

The above-described method and application are implemented as an open-source \textbf{R} package, \textbf{cheem} available on \href{https://CRAN.R-project.org/package=cheem}{CRAN}. Preprocessing was facilitated with models created via \textbf{randomForest} (Liaw and Wiener 2002) and explanations calculated with \textbf{treeshap} (Kominsarczyk et al. 2021). The application was made with \textbf{shiny} (Chang et al. 2021). The tour visual is built with \textbf{spinifex} (Spyrison and Cook 2020). Both views are created first with first with \textbf{ggplot2} (Wickham 2016) and then rendered as interactive HTML widgets with \textbf{plotly} (Sievert 2020). \textbf{DALEX} (Biecek 2018) and the free ebook, \emph{Explanatory Model Analysis} (Biecek and Burzykowski 2021) were a huge boon to understanding local explanations and how to apply them.

@@

To process your own data, you will need to use the \texttt{treeshap} package, which can be installed from GitHub:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{remotes}\SpecialCharTok{::}\FunctionTok{install\_github}\NormalTok{(}\StringTok{\textquotesingle{}ModelOriented/treeshap\textquotesingle{}}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Follow the examples provided with the package to compute the local explainers, (see \texttt{?cheem\_ls}). The application expects the return of call from \texttt{cheem\_ls()} which is then saved to an .rds file with \texttt{saveRDS()}. Alternatively, the cheem viewer shiny app can be directly accessed at \href{https://ebsmonash.shinyapps.io/cheem_initial/}{ebsmonash.shinyapps.io/cheem\_initial/}.

@@

The Palmer penguins data (Gorman, Williams, and Fraser 2014; Horst, Hill, and Gorman 2020) was collected on three species of penguins foraging near Palmer Station, Antarctica. The data was publicly available to substitute for the overly-used iris data and is quite similar in form. After removing incomplete instances, there are 333 instances, and we will use the four physical measurements, bill length (\texttt{b\_l}), bill depth (\texttt{b\_d}), flipper length (\texttt{f\_l}), body mass (\texttt{wgt}), for this illustration. A random forest model was fit with species as the response feature.

@@

\caption{Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is a Gentoo (purple) Chinstrap (orange) penguin that is misclassified as a Chinstrap (orange), marked as an asterisk in (a), and the dashed vertical line in (b). The radial view shows varying the contribution of \texttt{f\_l} from the initial attribution projection (b, left), which produces a linear combination where the PI is more probably (higher density value) a Chinstrap than a Gentoo (b, right). (The animation of the radial tour is at \href{https://vimeo.com/666431172}{vimeo.com/666431172}.)}\label{fig:casepenguins}
\end{figure}

@@

{\centering \includegraphics[width=0.6\linewidth]{./figures/case_penguins_BlFl}

@@

\caption{Checking what is learned from the cheem viewer. This is a plot of flipper length (\texttt{f\_l}) and bill length (\texttt{b\_l}), where an asterisk highlights the PI. A Gentoo (purple) misclassified as a Chinstrap (orange). The PI has an unusually small \texttt{f\_l} length which is why it is confused with a Chinstrap.}\label{fig:casepenguinsblfl}
\end{figure}

There is more separation visible in the attribution space than the data space, as would be expected. The predicted vs observed plot reveals a handful of misclassified instances. We will explore why a Gentoo has been wrongly labeled as a Chinstrap for this illustration. The PI is a misclassified point (represented by the asterisk in the global view and a dashed vertical line in the tour view). The CI is a correctly classified point (represented by an \(\times\) and a vertical dotted line).

The radial tour starts from the attribution projection of the misclassified instance (b, left). The important features identified by SHAP in the (wrong) prediction for this instance are mostly \texttt{b\_l} and \texttt{b\_d} with small contributions of \texttt{f\_l} and \texttt{wgt}. This projection is a view where the Gentoo (purple) looks much more likely for this instance than Chinstrap. That is, this combination of features is not particularly useful because the PI looks very much like other Gentoo penguins. To explore this, we use the radial tour to vary the contribution of flipper length (\texttt{f\_l}). (In our exploration, this was the third feature explored. It is typically useful to explore the features with larger contributions, here \texttt{b\_l} and \texttt{b\_d}, but when doing this, nothing was revealed about how the PI differed from other Gentoos). On varying \texttt{f\_l} as it contributes more to the projection (b, right), we see that more, and more, this penguin looks like a Chinstrap. This suggests that \texttt{f\_l} should be considered an important feature for explaining the (wrong) prediction.

Figure \ref{fig:casepenguinsblfl} confirms that flipper length (\texttt{f\_l}) is important for the confusion of the PI as a Chinstrap. Here, flipper length and body length are plotted, and we can see that the PI is closer to the Chinstrap group in these two features, mostly because it has an unusually low value of flipper length relative to other Gentoos. From this view, it makes sense that its a hard instance to account for as decision trees can only partition only vertical and horizontal lines.

@@

The chocolates dataset consists of 88 instances of ten nutritional measurements determined from their labels and labeled as either milk or dark. Dark chocolate is considered healthier than milk. The data was collected by students during the Iowa State University class STAT503 from nutritional information from the manufacturer's website and normalized to 100g equivalents. The data is available in the \textbf{cheem} package. A random forest model is used for the classification of chocolate type.

>>>>

\begin{document}
\maketitle
\begin{abstract}
The increased predictive power of nonlinear models comes at the cost of interpretability of its terms. This trade-off has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models use predictors to arrive at a prediction with \emph{local explanations}, a point estimate of the linear feature importance in the vicinity of one instance. These can be considered linear projections and can be further explored to understand better the interactions between features used to make predictions across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any instance and illustrate with examples with categorical (penguin species, chocolate types) and quantitative (soccer/football salaries, house prices) output. The methods are implemented in the \textbf{R} package \textbf{cheem}, available on CRAN.
\end{abstract}

@@

The remainder of this paper is organized as follows. The following Section, \ref{sec:explanations}, covers the background of the local explanation and the traditional visuals produced. Section \ref{sec:tour} explains the animations of continuous linear projections. Section \ref{sec:cheemviewer} discusses the visual layout in the graphical user interface, how it facilitates analysis, data preprocessing, and package infrastructure. Then Section \ref{sec:casestudies} illustrates the application to supervised learning with categorical and quantitative output. Section \ref{sec:cheemdiscussion} concludes with the insights gained and possible directions to explore in the future.

@@

SHaply Additive exPlanations (SHAP) quantifies the feature contributions of one instance by examining the effect of other features on the predictions. The explanations of SHAP refer to Shapley (1953)'s method to evaluate an individual's contribution to cooperative games by assessing this player's performance in the presence or absence of other players. Strumbelj and Kononenko (2010) introduced SHAP for local explanations in ML models. The attribution of feature importance depends on the sequence of the included features. The SHAP values are the mean contributions over different feature sequences. The approach is related to partial dependence plots (Molnar 2020), used to explain the effect of a feature by predicting the response for a range of values on this feature after fixing the value of all other features to their mean. Though partial dependence plots are a global approximation of the feature importance, while SHAP is specific to one instance. It could also be considered similar to examining the coefficients from all subsets regression, as described in Wickham, Cook, and Hofmann (2015), helps to understand the relative importance of each feature in the context of all other candidate features.

@@

\caption{Illustration of SHAP values for a random forest model FIFA 2020 player wages from nine skill predictors. A star offensive and defensive player are compared, L. Messi and V. van Dijk, respectively. Panel (a) shows breakdown plots of three sequences of the features. The sequence of the variables impacts the magnitude of their attribution. Panel (b) shows the distribution of attribution for each feature across 25 sequences of predictors, with the mean displayed as a dot for each player. Reaction skills are important for both players. Offense and movement are important for Messi but not van Dijk, and conversely, defense and power are important for van Dijk but not Messi.}\label{fig:shapdistrbd}
\end{figure}

Following the use case \emph{Explanatory Model Analysis} (Biecek and Burzykowski 2021), FIFA data is used to illustrate SHAP. Consider soccer data from the FIFA 2020 season (Leone 2020). There are 5000 instances of 9 skill measures (after aggregating highly correlated features). A random forest model is fit regressing player's wages {[}2020 Euros{]} from their skill measurements. The SHAP values are compared for a star offensive player (L. Messi) and defensive player (V. van Dijk). The results are displayed in Figure \ref{fig:shapdistrbd}. A difference in the attribution of the feature importance across the two positions of the players can be expected. This would be interpreted as how a player's salary depends on this combination of skill sets. Panel (a) is a modified breakdown plot (Gosiewska and Biecek 2019) where three sequences of features are presented, so the two instances can be more easily compared. The magnitude of the contributions depends on the sequence in which they appear. Panel (b) shows the differences of the player's median values of over 25 such sequences. In summary, these plots highlight how local explanations bring interpretability to a model, at least in the vicinity of their instances. In this instance, two players with different positions receive different profiles of feature importance to explain the prediction of their wages.

For the application, we use \emph{tree SHAP}, a variant of SHAP that enjoys a lower computational complexity (Lundberg, Erion, and Lee 2018). Instead of aggregating over sequences of the features, tree SHAP calculates instance-level feature importance by exploring the structure of the decision trees. Tree SHAP is only compatible with tree-based models; random forests are used for illustration. The following section will use normalized SHAP values as a projection basis (call this the \emph{attribution projection}) that will be used to explore the sensitivity of the feature contributions.

@@

To explore the local explanations, coordinated views (\textbf{roberts\_state\_2007?}) (also known as ensemble graphics, Unwin and Valero-Mora 2018) are provided in the \emph{cheem viewer} application. There are two primary plots: the \textbf{global view} to give the context of all of the SHAP values and the \textbf{radial tour view} to explore the local explanations with user-controlled rotation. There are numerous user inputs, including feature selection for the radial tour and instance selection for making comparisons. There are different plots used for the categorical and quantitative responses. Figures \ref{fig:classificationcase} and \ref{fig:regressioncase} are screenshots showing the cheem viewer for the two primary tasks: classification (categorical response) and regression (quantitative response).

@@

The global view provides a context of all instances and facilitates the exploration of the separability of the data- and attribution-spaces. Both of these spaces are of dimension \(n\times p\), where \(n\) is the number of instances and \(p\) is the number of predictors. The attribution space corresponds to the local explanations for each instance, each a vector of \(p\) values.

Visualization is provided by the first two principal components of the data (left) and the attribution (middle) spaces. These single 2D projections will not reveal all of the structure of higher-dimensional space, but they are helpful visual summaries. In addition, a plot of the observed against predicted response values is also provided (Figures \ref{fig:classificationcase}b, \ref{fig:regressioncase}a) to help identify instances poorly predicted by the model. For classification tasks, misclassified instances are circled in red. Linked brushing between the plots is provided, and a tabular display of selected points helps to facilitate exploration of the spaces and the model (shown in Figures \ref{fig:regressioncase}d).

While the comparison of these spaces is interesting, the primary purpose of the global view is to enable the selection of instances to explore the local explanations. The projection attribution of the primary instance (PI) is examined and typically viewed with an optional comparison instance (CI). These instances are highlighted as asterisk and \(\times\), respectively.

@@

The local explanations for all observations are normalized (squared sum of values adds to 1), and thus, the relative importance of features can be compared across all instances. These are depicted as vertical parallel coordinate plots (Ocagne 1885) on the basis biplot (Gabriel 1971). 1D biplot display depicts the values of the current basis as bars. The parallel coordinate overlays lines connecting one instance's feature attribution (Figures \ref{fig:classificationcase}e and \ref{fig:regressioncase}e). The attribution projections of the PI and CI are shown as dashed and dotted lines. From this plot, the range and density of the importance across all instances can be interpreted. For classification, one would look at differences between groups on any feature. For example, Figure \ref{fig:classificationcase}e suggests that \texttt{bl} is important for distinguishing the green class from the other two. For regression, one might generally observe which features have low values for all instances (not important), for example, \texttt{BMI} and \texttt{pwr} in Figure \ref{fig:regressioncase}e, and which have a range of high and low values (e.g.~\texttt{off}, \texttt{def}) suggesting they are important for some instances and not important for other instances.

The overlaid bars on the parallel coordinate plot represent the attribution projection of the PI. (Remember that the PI is interactively selected from the global view). The attribution projection is approximates the feature importance for predicting this instance. This combination of features best explains the difference between the mean response and an instance's predicted value. It is not an indication of the local shape of the model surface. That is, it is not some indication of the tangent to the curve at this point.

The attribution projection of the PI is the initial 1D basis in a radial tour, displayed as a density plot for a categorical response (Figure \ref{fig:classificationcase}f) and as scatterplots for a quantitative response (Figure \ref{fig:regressioncase}f). The PI and CI are indicated by vertical dashed and dotted lines, respectively. The radial tour varies the contribution of the selected feature between 0-1. This is viewed as an animation of the projections from many intermediate bases. Doing so tests the sensitivity of structure (class separation or strength of relationship) to the feature's contribution. For classification, if the separation between classes diminishes when the feature contribution is reduced, this suggests that the feature is important for class separation. For regression, if the relationship scatterplot weakens when the feature contribution is reduced, indicating that the feature is important for accurately predicting the response.

@@

Selecting a misclassified instance as PI and a correctly classified point nearby in data space as CI makes it easier to examine the features most responsible for the error. The global view (Figure \ref{fig:classificationcase}c) displays the model confusion matrix. The radial tour is 1D and displays as density where color indicates class. An animation slider enables the user to vary the contribution of each feature to explore the sensitivity of the separation to that feature.

@@

\caption{Overview of the cheem viewer for classification tasks. Global view inputs, (a), set the PI, CI, and color statistic. Global view, (b) PC1 by PC2 approximations of the data space and attribution space. (c) prediction by observed y (visual of the confusion matrix for classification tasks). Points are colored by predicted class, and red circles indicate misclassified instances. Radial tour inputs (d) select features to include and which feature is changed in the tour. (e) shows a parallel coordinate display of the distribution of the feature attributions while bars depict contribution for the current basis. The black bar is the variable being changed in the radial tour. Panel (f) is the resulting data projection indicated as density in the classification case.}\label{fig:classificationcase}
\end{figure}

@@

Selecting an inaccurately predicted instance as PI and an accurately predicted instance with similar feature values as CI is a helpful way to understand how the model is failing or not. The global view (Figure \ref{fig:regressioncase}a) shows a scatterplot of the observed vs predicted values, which should exhibit a strong relationship if the model is a good fit. The points can be colored by a statistic, residual, a measure of outlyingness (log Mahalanobis distance), or correlation aid in understanding the structure identified in these spaces.

In the radial tour view, the observed response and the residuals (vertical) are plotted against the attribution projection of the PI (horizontal). The attribution projection can be interpreted similarly to the predicted value from the global view plot. It represents a linear combination of the features, and a good fit would be indicated when there is a strong relationship with the observed values. This can be viewed as a local linear approximation if the fitted model is nonlinear. As the contribution of a feature is varied, if the value of the PI does not change much, it would indicate that the prediction for this instance is NOT sensitive to that feature. Conversely, if the predicted value varies substantially, the prediction is very sensitive to that feature, suggesting that the feature is very important for the PI's prediction.

@@

\caption{Overview of the cheem viewer for regression  and illustration of interactive features. Panel (a) PCA of the data and attributions spaces, (b) residual plot, predictions by observed values. Four selected points are highlighted in the PC spaces and tabularly displayed. Coloring on a statistic (c) highlights structure organized in the attribution space. Interactive tabular display (d) populates when instances are selected. Contribution of the 1D basis affecting the horizontal position (e) parallel coordinate display of the feature attribution from all observations, and horizontal bars show the contribution to the current basis. Regression projection (f) uses the same horizontal projection and fixes the vertical positions to the observed y and residuals (middle and right).}\label{fig:regressioncase}
\end{figure}

@@

The application has several reactive inputs that affect the data used, aesthetic display, and tour manipulation. These reactive inputs make the software flexible and extensible (Figure \ref{fig:classificationcase}a \& d). The application also has more exploratory interactions to help link points across displays, reveal structures found in different spaces, and access the original data.

A tooltip displays instance number/name and classification information while the cursor hovers over a point. Linked brushing allows the selection of points (left click and drag) where those points will be highlighted across plots (Figure \ref{fig:classificationcase}a \& b). The information corresponding to the selected points is populated on a dynamic table (Figure \ref{fig:classificationcase}d). These interactions aid exploration of the spaces and, finally, the identification of primary and comparison instances.

@@

It is vital to mitigate the render time of visuals, especially when users may want to iterate many explorations. All computational operations should be prepared before runtime. The work remaining when an application is run solely reacts to inputs and rendering visuals and tables. Below discusses the steps and details of the reprocessing.

\begin{itemize}
    \item \textbf{Data:} predictors and response are unscaled complete numerical matrix. Most models and local explanations are scale-invariant.
    \item \textbf{Model:} any model and compatible explanation could be explored with this method. Currently, random forest models are applied via the package \textbf{randomForest} (Liaw and Wiener 2002), compatibility tree SHAP. Modest hyperparameters are used, namely: 125 trees, number features randomly sampled at each split, mtry = $\sqrt{p}$ or $p/3$ for classification and regression, and minimum size of terminal nodes $max(1, n/500)$ or $max(5, n/500)$ for classification and regression. 
    \item \textbf{Local explanation:} Tree SHAP is calculated for \emph{each} instance using the package \textbf{treeshap} (Kominsarczyk et al. 2021). This implementation aggregates exhaustively over all trees' attribution and does not fit interactions of features.
    \item \textbf{Cheem viewer:} after the model and full explanation space are calculated, each feature is scaled by standard deviations away from the mean to achieve common support for visuals. Statistics for mapping to color are computed on the scaled spaces.
\end{itemize}

The time to preprocess the data will vary significantly with the complexity of the model and local explanation. For reference, the FIFA data contained 5000 instances of nine explanatory features took 2.5 seconds to fit a random forest model of modest hyperparameters. Extracting the tree SHAP values of each instance took 270 seconds total. PCA and statistics of the features and attributions took 2.8 seconds. These runtimes were from a non-parallelized R session on a modern laptop, but suffice to say that most of the time will be spent on the local attribution. An increase in model complexity or data dimensionality will quickly become an obstacle. Its reduced computational complexity makes tree SHAP an excellent candidate to start. (Alternatively, the package \textbf{fastshap} (Greenwell 2020) claims extremely low runtimes, attributed to fewer calls to the prediction function, partial implementation in C++, and efficient use of logical subsetting.)

@@

The above-described method and application are implemented as an open-source \textbf{R} package, \textbf{cheem} available on \href{https://CRAN.R-project.org/package=cheem}{CRAN}. Preprocessing was facilitated with models created via \textbf{randomForest} (Liaw and Wiener 2002) and explanations calculated with \textbf{treeshap} (Kominsarczyk et al. 2021). The application was made with \textbf{shiny} (Chang et al. 2021). The tour visual is built with \textbf{spinifex} (Spyrison and Cook 2020). Both views are created first with \textbf{ggplot2} (Wickham 2016) and then rendered as interactive \texttt{html} widgets with \textbf{plotly} (Sievert 2020). \textbf{DALEX} (Biecek 2018) and the free ebook, \emph{Explanatory Model Analysis} (Biecek and Burzykowski 2021) were a boon to understanding local explanations and how to apply them.

@@

Follow the examples provided with the package to compute the local explainers (see \texttt{?cheem\_ls}). The application uses the returned objects from a from \texttt{cheem\_ls()} call. Alternatively, the cheem viewer shiny app can be directly accessed at \href{https://ebsmonash.shinyapps.io/cheem_initial/}{ebsmonash.shinyapps.io/cheem\_initial/}.

@@

The Palmer penguins data (Gorman, Williams, and Fraser 2014; Horst, Hill, and Gorman 2020) was collected on three species of penguins foraging near Palmer Station, Antarctica. The data was publicly available to substitute for the overly-used iris data and is quite similar in form. After removing incomplete instances, there are 333 instances of four physical measurements, bill length (\texttt{bl}), bill depth (\texttt{bd}), flipper length (\texttt{fl}), body mass (\texttt{bm}), for this illustration. A random forest model was fit with species as the response feature.

@@

\caption{Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is a Gentoo (purple) penguin that is misclassified as a Chinstrap (orange), marked as an asterisk in (a), and the dashed vertical line in (b). The radial view shows varying the contribution of \texttt{fl} from the initial attribution projection (b, left), which produces a linear combination where the PI is more probably (higher density value) a Chinstrap than a Gentoo (b, right). (The animation of the radial tour is at \href{https://vimeo.com/666431172}{vimeo.com/666431172}.)}\label{fig:casepenguins}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{./figures/case_penguins_BlFl}

@@

\caption{Checking what is learned from the cheem viewer. This is a plot of flipper length (\texttt{fl}) and bill length (\texttt{bl}), where an asterisk highlights the PI. A Gentoo (purple) misclassified as a Chinstrap (orange). The PI has an unusually small \texttt{fl} length which is why it is confused with a Chinstrap.}\label{fig:casepenguinsblfl}
\end{figure}

There is more separation visible in the attribution space than the data space, as would be expected. The predicted vs observed plot reveals a handful of misclassified instances. A Gentoo that has been wrongly labeled as a Chinstrap is selected for illustration. The PI is a misclassified point (represented by the asterisk in the global view and a dashed vertical line in the tour view). The CI is a correctly classified point (represented by an \(\times\) and a vertical dotted line).

The radial tour starts from the attribution projection of the misclassified instance (b, left). The important features identified by SHAP in the (wrong) prediction for this instance are mostly \texttt{bl} and \texttt{bd} with small contributions of \texttt{fl} and \texttt{bm}. This projection is a view where the Gentoo (purple) looks much more likely for this instance than Chinstrap. That is, this combination of features is not particularly useful because the PI looks very much like other Gentoo penguins. To explore this, the radial tour is used to vary the contribution of flipper length (\texttt{fl}). (In our exploration, this was the third feature explored. It is typically helpful to explore the features with more significant contributions, here \texttt{bl} and \texttt{bd}. Still, when doing this, nothing was revealed about how the PI differed from other Gentoos). On varying \texttt{fl} as it contributes increasingly to the projection (b, right), more and more, this penguin looks like a Chinstrap. This suggests that \texttt{fl} should be considered an important feature for explaining the (wrong) prediction.

Figure \ref{fig:casepenguinsblfl} confirms that flipper length (\texttt{fl}) is vital for the confusion of the PI as a Chinstrap. Here, flipper length and body length are plotted, and the PI can be seen to be closer to the Chinstrap group in these two features, mainly because it has an unusually low value of flipper length relative to other Gentoos. From this view, it makes sense that it is a hard instance to account for as decision trees can only partition only vertical and horizontal lines.

@@

The chocolates dataset consists of 88 instances of ten nutritional measurements determined from their labels and labeled as either milk or dark. Dark chocolate is considered healthier than milk. The data were collected by students during the Iowa State University class STAT503 from nutritional information from the manufacturer's website and normalized to 100g equivalents. The data is available in the \textbf{cheem} package. A random forest model is used for the classification of chocolate types.

It could be interesting to examine the nutritional properties of any dark chocolates that have been misclassified as milk. A reason to do this is that a dark chocolate, nutritionally more like milk, should not be considered a healthy alternative. It is interesting to explore which nutritional features contribute most to misclassification.
